Patterns and anti-patterns

### 2021-03-24 analysis/patterns2.Rmd
Patterns

---
title: "Normalized expressions"
output:
    html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: Data/package/
  calls_path: summarized.fst
  normalized_path: normalized-expressions.csv
  corpus_path: corpus.txt
  static_path: evals-static.csv
  force_rebuild: FALSE
---

```{r setup, include=FALSE}
library(tidyverse)
library(fst)
library(fs)
library(DT)
library(scales)
library(corrr)
library(boot)

@@

# Tags will be prefixed by the dataset name
create_tags(path(TAGS_DIR, paste0(dataset_name, "_normalized_expr.tex")), prefix = dataset_name, default = TRUE)
# Also use ggsave_prefix instead of ggsave if you need to save a graph

@@

lock_path <- paste0(params$base_dir, ".normalized-lock.fst")
min_path <- paste0(params$base_dir, "Ecano.fst")
```

@@

# Ranking of normalized and minimized expressions

Normalized: 
```{r}
widespread_canonic <- E_cano %>%
  group_by(normalized) %>%
  summarize(n_call_sites = n_distinct(src_ref), n_packages = n_distinct(ev_package))

widespread_canonic %>%
  arrange(desc(n_call_sites)) %>%
  datatable()
```

Minimized:
```{r}
widespread_minimized <- E_cano %>%
  group_by(minimized) %>%
  summarize(n_call_sites = n_distinct(src_ref), proportion = ratio(n_call_sites, nb_call_sites), n_packages = n_distinct(ev_package), n_operations = round(mean(n_op, na.rm=TRUE), 2), percent_envir = ratio(n_distinct(cur_data()$src_ref[!is.na(cur_data()$envir)]),  n_call_sites), percent_envir_calls = ratio(sum(!is.na(envir)), n())) %>% 
  arrange(desc(n_call_sites))

widespread_minimized %>%
  datatable()
```

```{r}
nrows <- 12
r_vec("MinimizedCallSites", nrows, pull(widespread_minimized, n_call_sites))
r_vec("MinimizedPropSites", nrows, pull(widespread_minimized, proportion))
r_vec("MinimizedPackage", nrows, pull(widespread_minimized, n_packages))
r_vec("MinimizedOperations", nrows, pull(widespread_minimized, n_operations))
r_vec("MinimizedPercentEnvir", nrows, pull(widespread_minimized, percent_envir))
```

`n_operations` is the average of R interpreter eval loop per minimized expression. `prop_envir` shows how much often an *explicit* environment is passed to `eval`. Looking up to a variable and applying a function to a variable, are the two non-trivial categories that most of the time explicitly use the `envir` argument of `eval`.

# How does it evolve per call site

To check whether the normalization makes sense, we have a look at how many normalized expressions we have per call sites.

```{r}
# count hash because resolved expressions are trimmed but hash are on the full expression
by_src_ref <- E_cano %>% group_by(src_ref) %>%
  summarise(n_resolved = n_distinct(hash), n_normalized = n_distinct(normalized), n_minimized = n_distinct(minimized))
```

```{r}
by_src_ref %>% arrange(desc(n_resolved)) %>% datatable()
```

There should always be less normalized expressions than resolved ones. 
We can see for how many there are strictly less (or one when it was already one).

```{r}
decrease_exprs <- by_src_ref %>% filter(n_resolved == 1 | n_resolved > n_minimized)  %>% arrange(desc(n_minimized))
```

It corresponds to `r nrow(decrease_exprs)` call sites so only `r nb_call_sites - nrow(decrease_exprs)` have non-trivially not decreased the number of call sites.

# Top calls

For the functions at the root of the AST, per call site:

```{r}
widespread_topcall <- E_cano %>%
  group_by(topcall) %>%
  summarize(n_call_sites = n_distinct(src_ref), percent = round(ratio(n_call_sites, nb_call_sites), 2)) %>% 
  arrange(desc(n_call_sites))
```

```{r}
widespread_topcall %>% datatable()
```

```{r}
big_topcall <- widespread_topcall %>% filter(n_call_sites > 3) 
```

There are `r nrow(big_topcall)` top calls out of `r nrow(widespread_topcall)` that are present in strictly more than 3 call sites.

# Variables

The percentage of non-default envir for variables seem to be quite low for the use-case. What if we remove more complex expressions and just keep the symbols?

```{r}
var_minimized <- E_cano %>% filter(minimized == "X")
nb_var_min_sites <- var_minimized %>% pull(src_ref) %>% n_distinct()

var_symb <-  E_cano %>% filter(minimized == "X", type == "symbol")
nb_var_symb_sites <- var_symb %>% pull(src_ref) %>% n_distinct()

var_symb_non_default_envir <- var_symb %>% filter(!is.na(envir))
nb_var_symb_ndefault_sites <- var_symb_non_default_envir %>% pull(src_ref) %>% n_distinct()

r("nb non default envir variables percent", ratio(nb_var_symb_ndefault_sites, nb_var_symb_sites))
```

What happens for the symbol with the default envir?

```{r}
var_sym_default_envir <- var_symb %>% filter(is.na(envir)) %>% count(src_ref, sort = TRUE)
```

Seems it was a bug: when the explicit argument was `parent.frame()`, it was counted as being the default argument (but it does not have the same semantics as a default argument or as an explicit one though!).

# Function definitions

```{r}
fundefs <- E_cano %>% filter(has_fundef)

nb_fundef_callsites <- fundefs %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_fundef_callsites` call sites with function definitions, i.e. `r ratio(nb_fundef_callsites, nb_call_sites)` %.

# Model.frame

```{r}
modelframes <- E_cano %>% filter(is_model)

nb_modelf_call_sites <- modelframes %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_modelf_call_sites` call sites with `mode.frame`, i.e. `r ratio(nb_modelf_call_sites, nb_call_sites)`%.

Is it usually in top call position?

```{r}
modelframes %>% filter(topcall != "model.frame")
```

Yes, always!

Is it correlated with `match.call`?

```{r}
nb_mf_match.call <- modelframes %>% filter(!is.na(match.call)) %>% pull(src_ref) %>% n_distinct()
```

`r nb_mf_match.call` call sites out of `model.frame` `r nb_modelf_call_sites` call sites use the `match.call` pattern, i.e. `r ratio(nb_mf_match.call, nb_modelf_call_sites)`%.

We can also reciprocally look at how many call sites with `match.call` have `model.frame`:

```{r}
match.call_df <- E_cano %>% filter(!is.na(match.call))

nb_m.c_call_sites <- match.call_df %>% pull(src_ref) %>% n_distinct()

nb_m.c_mf_call_sites <- match.call_df %>% filter(is_model) %>% pull(src_ref) %>% n_distinct()
```

So `r ratio(nb_m.c_mf_call_sites, nb_m.c_call_sites)`% of the call sites with the `match.call` pattern use `model.frame`.

# Assignments

When do we perform assignments (causing potentially side effects)?

```{r}
assigns <- E_cano %>% filter(has_assigns > 0)

nb_assigns_call_sites <- assigns %>% pull(src_ref) %>% n_distinct()

r("nb assign sites", nb_assigns_call_sites)
r("assign sites percent", ratio(nb_assigns_call_sites, nb_call_sites))
```

There are `r nb_assigns_call_sites` call sites with assignments, i.e. `r ratio(nb_assigns_call_sites, nb_call_sites)`%.

Is the `envir` argument non trivial?

```{r}
assign_env <- assigns %>% filter(!is.na(envir))

nb_a_env_call_sites <- assign_env %>% pull(src_ref) %>% n_distinct()

r("non default envir assign sites", nb_a_env_call_sites)
r("non default envir assign sites percent", ratio(nb_a_env_call_sites, nb_assigns_call_sites))
```

There are `r nb_a_env_call_sites` call sites where `envir` is non-trivial, with an assignment, i.e. `r ratio(nb_a_env_call_sites, nb_assigns_call_sites)`% of the call sites with assignments.

# `match.call` pattern

There are `r nb_m.c_call_sites` call sites using the `match.call`pattern, i.e. `r ratio(nb_m.c_call_sites, nb_call_sites)`%.

The top calls for this pattern are:

@@

Most are `model.frame`. If we add other model functions, such as `glm` and `plm`:

@@

it amounts for `r nb_stat_call_sites` call sites, i.e `r ratio(nb_stat_call_sites, nb_call_sites)`%.

# Substitute&co

`substitute`, `quote` and so on are used to do meta-programming.

```{r}
meta_df <- E_cano %>% filter(has_meta_op)

nb_meta_call_sites <- meta_df %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_meta_call_sites` call sites with `substitute` and `substitute`-like operators.

Are the `substitute` sometimes not placed as a top call?

```{r}
meta_not_top <- meta_df %>% filter(!topcall %in% c("substitute", "quote", "enquote", "bquote"))

nb_m_n_top_call_sites <- meta_not_top %>% pull(src_ref) %>% n_distinct()
```

It happens for `r nb_m_n_top_call_sites` call sites, i.e. `r ratio(nb_m_n_top_call_sites, nb_call_sites)`%.

@@

There are `r nb_plotting_call_sites` plotting call sites, i.e. `r ratio(nb_plotting_call_sites, nb_call_sites)`%.

### Infix operators

@@

# Polymorphism in more details

We have only 15 different kind of minimized expression. Do some minimized expressions often go together?

```{r}
minimized_exprs <- E_cano %>% pull(minimized) %>% unique()

create_set <- function(l) {
  s <- rep.int("F", length(minimized_exprs))
  names(s) <- minimized_exprs

  for(exp in l) {
    s[[exp]] <- "T"
  }

  return(paste0(s, collapse = ""))
}

index_v <- match("V", minimized_exprs)
index_x <- match("X", minimized_exprs)

has_value <- function(l) {
  substr(l, index_v, index_v) == "T"
}

is_variable <- function(l) {
  substr(l, index_x, index_x) == "T"
}

count_elements <- function(l) {
  sum(map_int(strsplit(l, split="")[[1]], function(ch) {if(ch == "T") {1L} else {0L}}))
}

show_set <- function(s) {
  l <- strsplit(s, split="")[[1]]
  names(l) <- minimized_exprs

  l <- l[which(l == "T")]
  return(paste0(names(l), collapse = "; "))
}

poly <- E_cano %>% group_by(src_ref) %>%
    summarize(common_minimized = create_set(minimized), n_exprs = n_distinct(minimized)) 
```

Now, we can show the mot common minimized expressions when there are more than 2 for one call site:

```{r}
poly %>%
  count(common_minimized,  sort = TRUE) %>%
  mutate(common_minimized = map_chr(common_minimized, show_set)) %>% 
  mutate(percent = round(100 * n / nb_call_sites, 2)) %>% 
  datatable()
```

How many sites have only one minimized expression?

```{r}
n_one_mini <- poly %>% filter(n_exprs == 1) %>% nrow()

r("Nb One Minimized", n_one_mini)
r("Nb One Minimized Percent", ratio(n_one_mini, nb_call_sites))
```

# Values

Many call sites only evaluate a value. It is either a language expression or symbol, or an actual value.

```{r}
value_minimized <- widespread_minimized %>% filter(minimized == "V")
```

These are the number of call sites that are actual values.
```{r}
main_types <- E_cano %>% group_by(src_ref) %>% summarise(nb_types = n_distinct(type), type = first(type), nb_runs = n_distinct(file))

only_actual_values <- main_types %>% filter(!type %in% c("symbol", "language", "expression", "promise"), nb_types == 1)
nb_actual_values_sites <- only_actual_values %>% pull(src_ref) %>% n_distinct()

r("nb call sites unique actual value", nb_actual_values_sites)
r("call sites unique actual value percent", ratio(nb_actual_values_sites, value_minimized$n_call_sites))
r("median run sites unique actual value", median(only_actual_values$nb_runs, na.rm = TRUE))
r("average run sites unique actual value", mean(only_actual_values$nb_runs, na.rm = TRUE))
r("sd run sites unique actual value", sd(only_actual_values$nb_runs, na.rm = TRUE))
```

Is it a simple or a complex value?

```{r}
val_exprs <- E_cano %>% filter(minimized == "V") %>%
  mutate(caller_srcref = if_else(is.na(caller_srcref), file, caller_srcref)) %>%
  group_by(src_ref) %>%
  summarise(avg_ast_nodes = mean(ast_nodes, na.rm = TRUE), avg_arg_size = mean(arg_size, na.rm = TRUE), nb_runs = n_distinct(file), nb_caller_srcref = n_distinct(caller_srcref), nb_calls = sum(duplicates), nb_caller_exprs = n_distinct(caller_expression)) %>%
  arrange(desc(avg_ast_nodes))

n_val_exprs <- val_exprs %>% nrow()
n_one_node <- val_exprs %>% filter(avg_ast_nodes == 1) %>% nrow()

r("val one node", n_one_node)
r("val one node percent", ratio(n_one_node, n_val_exprs))
```

Are the sites resolving to just a value well covered?

```{r}
med_runs <- median(val_exprs$nb_runs, na.rm = TRUE)
avg_runs <- mean(val_exprs$nb_runs, na.rm = TRUE)
sd_runs <- sd(val_exprs$nb_runs, na.rm = TRUE)

r("one val median runs", med_runs)
r("one val average runs", avg_runs)
r("one val sd runs", sd_runs)
```

Is there a correlation between the number of caller srcref and the presence of other minimized expressions for one call site?

There are many `NA` caller srcref, so we count them as 1 currently...

```{r eval = FALSE}
val_exprs_comb <- val_exprs %>% left_join(main_types, by="src_ref")  %>%
  mutate(calls_per_run = nb_calls / nb_runs) %>%
  mutate(calls_per_site = nb_calls / nb_caller_srcref) %>%
  mutate(calls_per_site_per_run = nb_calls / nb_runs / nb_caller_srcref)
# normalization to be able to see both on the same graph

val_exprs_comb %>% ggplot(aes(x = nb_caller_srcref, y = nb_types)) + 
  geom_jitter(alpha = 0.3) +
  scale_x_log10()

val_exprs_comb %>% ggplot(aes(x = nb_runs, y = nb_types)) + 
  geom_jitter(alpha = 0.3) +
  scale_x_log10()

val_exprs_comb %>% ggplot(aes(x = calls_per_run, y = nb_types)) + 
  geom_jitter(alpha = 0.3) +
  scale_x_log10()

val_exprs_comb %>% ggplot(aes(x = calls_per_site, y = nb_types)) + 
  geom_jitter(alpha = 0.3) +
  scale_x_log10()

val_exprs_comb %>% ggplot(aes(x = calls_per_site_per_run, y = nb_types)) + 
  geom_jitter(alpha=0.3) +
  scale_x_log10()

val_exprs_comb %>% ggplot(aes(x = nb_caller_exprs, y = nb_types)) + 
  geom_jitter(alpha=0.3) +
  scale_x_log10()

df <- val_exprs_comb %>% select(where(is.numeric))

df %>% correlate(method = "spearman") %>% network_plot()
df %>% correlate(method = "spearman") %>% rplot() + theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=1))

Manual inspection of the source code shows that there are actually some call sites that take actual values.
If a site is exercised by many caller sites, or by many different runs, and has only values, then we reckon it is probably a _useless_ `eval`...

```{r}
val_only <- val_exprs %>% right_join(only_actual_values, by = c("src_ref", "nb_runs"))

max_val_runs <- val_only %>% filter(nb_caller_exprs == max(nb_caller_exprs)) #it happens to also be the one with the most caller expressions

val_only %>% filter(nb_caller_exprs > 5) # those ones are likely to be useless eval
```

And the results of the manual inspection:

```{r, include=FALSE}
useful_percent <- 0
```

```{r, eval=dataset_name == "package"}
manual_value_necessary <- read_csv(path(params$base_dir, "non-necessary-value-evals.csv"), col_types = cols(src_ref = col_character(), necessary = col_logical()))

statistic <- function(data, i) {
  df <- data[i,] %>% select(necessary) %>% filter(necessary)
  nrow(df) / nrow(data[i,])
}

bootstrap <- boot(manual_value_necessary, statistic, R = 1000)

#useful_percent <- mean(bootstrap$t, na.rm = TRUE)
useful_percent <- bootstrap$t0

r("useful value eval percent", percent(useful_percent))
```

`r percent(useful_percent)`% of those `eval` are useful.

We now want to characterize better the sites that are passed a non-default environment. Are they also the sites that can get a value or something with a variable inside?

We compute the sites which are at least one given a `V` minimized expression. Then we look at whether they are passed the default envir or not, and if there are variables in:

```{r}
minimized_sites <- E_cano %>% semi_join(val_exprs, by="src_ref") %>% group_by(src_ref) %>% 
  summarize(has_envir = all(!is.na(envir)), has_var = any(has_var, na.rm = TRUE))

non_default_envir <- minimized_sites %>% filter(has_envir)

non_default_with_var <- minimized_sites %>% filter(has_envir, has_var)

r("non default envir value percent", ratio(nrow(non_default_envir), nrow(minimized_sites)))
r("non default envir with var percent", ratio(nrow(non_default_with_var), nrow(non_default_envir)))
```

# Accessing a slot

Accessing a slot is done with `$`, `[]` or `[[]]`.

```{r}
slot_access <- E_cano %>% filter(minimized == "$")

generalized_slot_access <- E_cano %>% filter(has_dollar | has_bracket)

nb_slot_sites <- slot_access %>% pull(src_ref) %>% n_distinct()

nb_generalized_slot_sites <- generalized_slot_access %>% pull(src_ref) %>% n_distinct()

r("nb slot access", nb_slot_sites)
r("slot access percent", ratio(nb_slot_sites, nb_call_sites))
r("generalized slot access", nb_generalized_slot_sites)
r("generalized slot access percent", ratio(nb_generalized_slot_sites, nb_call_sites))
```

>>>>

---
title: "Patterns and anti-patterns"
output:
    html_document:
        gallery: false
        toc: true
        toc_depth: 3
        toc_float: true
        df_print: paged
date: "`r Sys.Date()`"
editor_options: 
  chunk_output_type: console
params:
  base_dir: Data/package/
  calls_path: summarized.fst
  normalized_path: normalized-expressions.csv
  corpus_path: corpus.txt
  static_path: evals-static.csv
  force_rebuild: FALSE
---

```{r setup, include=FALSE}
library(tidyverse)
library(fst)
library(fs)
library(DT)
library(scales)
library(corrr)

@@

# Tags will be prefixed by the dataset name
create_tags(path(TAGS_DIR, paste0(dataset_name, "_patterns.tex")), prefix = dataset_name, default = TRUE)
# Also use ggsave_prefix instead of ggsave if you need to save a graph

@@

lock_path <- paste0(params$base_dir, ".patterns-lock.fst")
min_path <- paste0(params$base_dir, "Epatterns.fst")
```

@@

# Variable lookup in another environment

Variables lookup are characterized by:

- symbol as the type of the expression
- a non-default environment

```{r}
variable_lookup <- E_cano %>% filter(type == "symbol", !is.na(envir))

nb_variable_lookup <- variable_lookup %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_variable_lookup` sites with variable lookup, i.e `r ratio(nb_variable_lookup, nb_call_sites)`%.

# `match.call`

The goal is to get a call and modify it or use it.

```{r}
match.call_df <- E_cano %>% filter(!is.na(match.call))
nb_m.c_call_sites <- match.call_df %>% pull(src_ref) %>% n_distinct()

There are `r nb_m.c_call_sites` sites using `match.call`, i.e. `r ratio(nb_m.c_call_sites, nb_call_sites)`%.

@@

## Evaluating some of the call arguments

In that case, `topcall` is `NA`, `$` or `[`.

```{r}
mc_arg_functions <- c(NA_character_, "$", "[")

nb_arg_mc_sites <- match.call_df %>% 
  filter(topcall %in% mc_arg_functions) %>% 
  pull(src_ref) %>% 
  n_distinct()
```

it amounts for `r nb_arg_mc_sites` call sites, i.e `r ratio(nb_arg_mc_sites, nb_m.c_call_sites)`%.

## `match.call` and statistical models

Veyr often, `match.call` is used to build statistical models, by replacing the call function by `model.frame` (but also other statistical models, such as `glm`).

@@

it amounts for `r nb_stat_call_sites` call sites, i.e `r ratio(nb_stat_call_sites, nb_m.c_call_sites)`%.

# Parsing strings

As a first approximation, it corresponds to type `expression` (although `expr` can also build such expressions). However, it is an under-approximation, because the expression could be later used...

```{r}
parse_eval <- E_cano %>% filter(type == "expressions" | str_detect(ev_expr, fixed("parse")))

nb_parse_eval <- parse_eval %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_parse_eval` sites using `parse`, i.e. `r ratio(nb_parse_eval, nb_call_sites)`%.

## Logging

https://github.com/cran/Rcmdr/blob/4b196764a44a047f8c7be3f4c629f52bee2c0c80/R/commander.R#L1301

The goal is to print the command and then to execute it.

## Formulas

```{r}
formulas <- E_cano %>% filter(str_detect(ev_expr, fixed("parse")), topcall == "~" | str_detect(ev_expr, fixed("formula")))

nb_formula_sites <- formulas %>% pull(src_ref) %>% n_distinct()
```

There are `r nb_formula_sites` sites building formulas with `parse`, i.e. `r ratio(nb_formula_sites, nb_call_sites)`%.

There are `r nb_plotting_call_sites` plotting call sites, i.e. `r ratio(nb_plotting_call_sites, nb_call_sites)`% and `r ratio(nb_plotting_call_sites, nb_parse_eval)`% of the parsing sites.

@@

# Symbolic computations

`D` operator to derive an expression symbolically, `Simplify`

For instance: https://github.com/cran/Deriv/blob/9ca174a959d8eb39b0a1f1a70de26d147a837176/R/Simplify.R#L572

Usually on an expression object, not a language object. Environment type seems to be `list` in that case, most often.

# Infix operators

@@

# Meta-programming

As a first approximation, it corresponds to type `language`.

```{r}
parse_eval <- E_cano %>% filter(type == "language")

nb_parse_eval <- parse_eval %>% pull(src_ref) %>% n_distinct()
```

## selecting columns in a dataframe

https://github.com/cran/epiDisplay/blob/d3d129a690a84b95c9aec437690227f1d50e3394/R/epiDisplay.R#L5915

```r 
 selected <- eval(substitute(vars), nl, parent.frame())
```

`vars` correspond to the columns in the dataframe we want to use.

Some examples on how it is called. Columns can be selected by name or by column number.

```r 
tableStack(bakedham:fruitsalad, .data)

tableStack(vars=4:25, .data, by=Origin)

tableStack(c(qa1,qa13:qa18,mean.score,total.score), .data, by=sex, test=FALSE)
```

# Non-necessary `eval`

## Values into `eval`
